### 2015-01-27 doc/thesis/system.tex
Improving the text about typing rules

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\vec{l} = el \; | \;
m \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
\mathbf{return} \; el \; | \;
e(el)_{s} \; | \;
e{:}n(el)_{s}\\
e ::= & \;\; \mathbf{nil} \; | \;
c \; | \;
{...}_{e} \; | \;
n_{e} \; | \;
e_{1}[e_{2}] \; | \;
{<}t{>} \; n \; | \;
f \; | \;
\{ \} \; | \;
\{ \; \vec{p} \; \} \; | \;
\{ \; {...}_{m} \; \} \; | \;
\{ \; \vec{p},{...}_{m} \; \}\\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
e(el)_{e} \; | \;
e{:}n(el)_{e}\\
l ::= & \;\; n_{l} \; | \;
e_{1}[e_{2}] \; | \;
n[c] \; {<}t{>}\\
c ::= & \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string}\\
p ::= & \;\; [e_{1}] = e_{2}\\
el ::= & \;\; \mathbf{nothing} \; | \;
\vec{e} \; | \;
me \; | \;
\vec{e}, me\\
me ::= & \;\; e(el)_{m} \; | \;
e{:}n(el)_{m} \; | \;
{...}_{m}\\
m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (){:}r \; fb \; | \;
\mathbf{fun} \; n_{1}{:}n_{2} \; ({...}{:}t){:}r \; fb\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; fb \; | \;
\mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
f ::= & \;\; \mathbf{fun} \; (){:}r \; fb \; | \;
\mathbf{fun} \; ({...}{:}t){:}r \; fb \; | \;
\mathbf{fun} \; (\vec{n{:}t}){:}r \; fb \; | \;
\mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
fb ::= & \;\; s \;;\; \mathbf{return} \; el\\
n ::= & \;\; {\it name}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}

@@

We present the typing rules as a deduction system for the typing
relations $\env_{1}, \penv \vdash s, \env_{2}$ and
$\env_{1}, \penv \vdash e : t, \env_{2}$.
We use the first one for typing statements, and the second one
for typing expressions.
The first relation means that given the type environment $\env_{1}$
and the projection environment $\penv$,
typing the statement $s$ produces the new type environment $\env_{2}$.
The second relation means that given the type environment $\env_{1}$
and the projection environment $\penv$,
the expression $e$ has type $t$ and produces the new type environment $\env_{2}$.
We need two typing relations because only expressions have types,
though expressions and statements can modify the environment.
We need two environments because $\env$ maps variables to types,
while $\penv$ maps projection variables to unions of tuple types.
We use $\env_{1}[n \mapsto t]$ for meaning that we extended the environment
$\env_{1}$ with the variable $n$ that maps to type $t$.
We use $\penv[x \mapsto r]$ for meaning that we extendend the environment
$\penv$ with the projection variable $x$ that maps to the union of tuple types $r$.

Lua has multiple assignments, and the rule \textsc{T-ASSIGNMENT} uses
tuple types along with consitent-subtyping to type check this Lua feature:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\begin{array}{c}
\dfrac{\env_{1}, \penv \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
       r_{1} \lesssim r_{2}}
      {\env_{1}, \penv \vdash \vec{l} = el,\env_{3}}
\end{array}
\end{array}
\]

@@

First, this rule uses the same environment $\env$ to type check each
left-hand side value $l_{k}$, as they can perform different changes
in the environment.
Then, it takes each type $t_{k}$ and builds the tuple type for
the assignment rule.
It places the type $\Value{*}$ in the end to discard extra values.
The rule succeeds if the predicate \emph{merge} can
produce a new environment, which includes all the modifications
of each environment $\env_{k}$.
Intuitively, the predicate \emph{merge} fails when an environment
tries to change the type of a variable towards a type that is not
a subtype of its previous type.
We will show in this section that the refinment of table types can
change the environment, and that is the reason why we need to check
whether the environment is consistent.

@@

When our type system type checks an expression list,
it always include $\Nil{*}$ in the end of the type of this expression list,
if the type does not end in a variadic type.
Our type system has this behavior because $\Nil{*}$ along with subtyping
and consistent-subtyping use $\Nil$ to replace missing values.
This behavior preserves the semantics of Lua on replacing missing values,
as it is necessary when we ommit optional parameters in a function call,
like the previous example showed.

@@

We do not allow the refinement of table types to change the type of
a field that is already present in the table type.
For instance,
\[
\string_ENV[``x"] \; {<}\String{>} = ``foo" \;;\; \string_ENV[``x"] \; {<}\Integer{>} = 1
\]
do not type check, as we are trying to add a field that already exists
in $\string_ENV$.
We could have used union types to allow the refinement of existing fields,
but this could lead to fields that have a too general type, which could
decrease the ammount of static type checking.
This means that our type system does not allow the table type to change
towards a type that is not a subtype of the previous type.
In this example,
\[
\{``x":\String \cup \Integer\}_{open} \not\subtype \{``x":\String\}_{open}
\]

@@

Sometimes we want to use a local variable that holds an \emph{unique}
table type to initialize a \emph{closed} table type, as \emph{unique}
table types allow us to initialize optional fields.
To do that, Typed Lua includes a cast expression that allow us
to use an \emph{unique} table type before it becomes \emph{open}.
The rule \textsc{T-CAST} describes this behavior:
\[
\begin{array}{c}
\mylabel{T-CAST}\\
\dfrac{\env_{1}(n) \subtype t \;\;\;
       \env_{1}[n \mapsto t], \penv \vdash n_{e}:t_{1}, \env_{2}}
      {\env_{1}, \penv \vdash {<}t{>} \; n:t_{1}, \env_{2}}
\end{array}
\]

@@

This rule prevents the following unsafe example to type check:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; b:\Integer \times \Integer \rightarrow \Integer =$}\\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer, y:\Integer):\Integer$}\\
& & & \multicolumn{1}{l}{$a[``z"] \; {<}\Integer{>} = 1; \; \mathbf{return} \; x + y$}\\
& \multicolumn{3}{l}{$\mathbf{in} \; a[``z"] \; {<}\String{>} = ``foo"$}
\end{tabular}
\end{center}

Typed Lua uses the refinement of table types to allow programmers to
build modules and objects.
It also allows Typed Lua to type check an object-oriented idiom that
programmers often use it.
The rule \textsc{T-METHOD3} illustrates how our type system type checks
method definitions:
\[
\begin{array}{c}
\mylabel{T-METHOD3}\\
\dfrac{\begin{array}{c}
       \env_{1}(n_{1}) = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{unique|open} \;\;\;
       \env_{1}, \penv \vdash n_{2} : l \;\;\;
       \not \exists i \in 1..n \; l \lesssim k_{i}\\
       closeall(\env_{1}[self \mapsto \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{closed}, \vec{n} \mapsto \vec{t}, \ret \mapsto r]), \penv \vdash s, \env_{2}\\
       t_{m} = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}, l{:}\Const \; \Self \times \vec{t} \rightarrow r\}_{unique|open}
       \end{array}}
      {\begin{array}{c}
       \env_{1}, \penv \vdash \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; s,\\
       closeset(\env_{1}[n_{1} \mapsto t_{m}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))
       \end{array}}
\end{array}
\]

The rule \textsc{T-METHOD3} allow our type system to type check the
following example, which is similar to the definition of the class
\texttt{Shape} from Section \ref{sec:oop}:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{open} = \{ [``x"] = 0.0, [``y"] = 0.0 \}$}\\
\multicolumn{4}{l}{$\mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{fun} \; s{:}new (x:\Number, y:\Number):\Self$}\\
& & \multicolumn{2}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{closed} =$}\\
& & & \multicolumn{1}{l}{$setmetatable(\{\}, \{ [``\string_\string_index"] = self \})$}\\
& & \multicolumn{2}{l}{$\mathbf{in} \; s[``x"] = x; \; s[``y"] = y; \; \mathbf{return} \; s$}\\
; & \multicolumn{3}{l}{$\mathbf{fun} \; s{:}move (x:\Number, y:\Number):\Void$}\\
& & \multicolumn{2}{l}{$self[``x"] = self[``x"] + x;$}\\
& & \multicolumn{2}{l}{$self[``y"] = self[``y"] + y$}
\end{tabular}
\end{center}

@@

Another common idiom that programmers use in Lua is to overload
the input parameter of functions, and use the function \texttt{type}
to execute different actions according to their types.
The rule \textsc{T-IF2} shows the case where our type system
discriminates the type of a local variable base on the tag \texttt{string}:
\[
\begin{array}{c}
\mylabel{T-IF2}\\
\dfrac{\begin{array}{c}
       \env_{1}(n) = t\\
       closeall(\env_{1}[n \mapsto \String]), \penv \vdash s_{1}, \env_{2} \\
       closeall(\env_{1}[n \mapsto filter(t, \String)), \penv \vdash s_{2}, \env_{3}\\
       \env_{4} = closeset(\env_{1}[n \mapsto t], fav(s_{1}) \cup fav(s_{2}))
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; type(n) == ``string" \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}

@@

Using the rule \textsc{T-IF2}, Typed Lua can type check the following example:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; o:\String \times \String \cup \Integer \rightarrow \String =$}\\
& \multicolumn{3}{l}{$\mathbf{fun} \; (a:\String, b:\String \cup \Integer):\String$}\\
& & \multicolumn{2}{l}{$\mathbf{local} \; r:\String = ``" \; \mathbf{in}$}\\
& & & \multicolumn{1}{l}{$\mathbf{if} \; type(b) == ``string" \; \mathbf{then} \; r = a \;{..}\;b \; \mathbf{else} \; r = rep(a, b)_{e}\; ;$}\\
& & & \multicolumn{1}{l}{$\mathbf{return} \; r$}\\
\multicolumn{4}{l}{$\mathbf{in} \; o(``foo", 2)_{s}$}
\end{tabular}
\end{center}

We are assuming that the function \emph{rep} is in the environment and has type
$\String \times \Integer \times \String \cup \Nil \times \Void \rightarrow \String \times \Void$.

Typed Lua also includes similar rules to handle the tags \texttt{nil},
\texttt{boolean}, and \texttt{number}.
There is also a similar rule for handling the type $\Integer$, but
it works only with Lua 5.3, as it depends on the function \texttt{math.type}.
This function appears only in Lua 5.3 and it returns the string $``integer"$
when its input parameter is a number that has an integer representation,
the string $``float"$ when its input parameter is a number that has a
floating point representation, or $\Nil$ otherwise.

Lua programmers also overload the return type of functions to denote
errors, and Typed Lua also handles this case with the rule \textsc{T-IF3}:
\[
\begin{array}{c}
\mylabel{T-IF3}\\
\dfrac{\begin{array}{c}
       \env_{1}(n) = \pi_{i}\\
       closeall(\env_{1}[\pi \mapsto fpt(\env_{1}(\pi), \Nil, i)]), \penv \vdash s_{1}, \env_{2} \\
       closeall(\env_{1}[\pi \mapsto gpt(\env_{1}(\pi), \Nil, i)), \penv \vdash s_{2}, \env_{3}
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; n \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2}))}
\end{array}
\]

\[
\begin{array}{c}
\mylabel{T-EXPLIST6}\\
\dfrac{\env, \penv \vdash me:t_{1} \times ... \times t_{n} \times \Void \sqcup t_{1}' \times ... \times t_{n}' \times \Void, \env_{1}}
      {\env, \penv \vdash me:\pi_{1} \times ... \times \pi_{n} \times \Nil{*}, \env_{1}}
\end{array}
\]

\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; q:\pi_{1}, r:\pi_{2} = idiv(1, 2)_{m} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; print(q + r)_{s} \; \mathbf{else} \; print(``ERROR: " \; .. \; r)_{s}$}
\end{tabular}
\end{center}

We are assuming that functions \emph{idiv} and \emph{print} are in
the environment with the respective types
$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void$
and
$\Value{*} \rightarrow \Void$.

>>>>

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\vec{l} = el \; | \;
m \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
\mathbf{return} \; el \; | \;
e(el)_{s} \; | \;
e{:}n(el)_{s}\\
e ::= & \;\; \mathbf{nil} \; | \;
c \; | \;
{...}_{e} \; | \;
n_{e} \; | \;
e_{1}[e_{2}] \; | \;
{<}t{>} \; n \; | \;
f \; | \;
\{ \} \; | \;
\{ \; \vec{p} \; \} \; | \;
\{ \; {...}_{m} \; \} \; | \;
\{ \; \vec{p},{...}_{m} \; \}\\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
e(el)_{e} \; | \;
e{:}n(el)_{e}\\
l ::= & \;\; n_{l} \; | \;
e_{1}[e_{2}] \; | \;
n[c] \; {<}t{>}\\
c ::= & \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string}\\
p ::= & \;\; [e_{1}] = e_{2}\\
el ::= & \;\; \mathbf{nothing} \; | \;
\vec{e} \; | \;
me \; | \;
\vec{e}, me \; | \;
me^{x} \; | \;
\vec{e}, me^{x}\\
me ::= & \;\; e(el)_{m} \; | \;
e{:}n(el)_{m} \; | \;
{...}_{m}\\
m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (){:}r \; fb \; | \;
\mathbf{fun} \; n_{1}{:}n_{2} \; ({...}{:}t){:}r \; fb\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; fb \; | \;
\mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
f ::= & \;\; \mathbf{fun} \; (){:}r \; fb \; | \;
\mathbf{fun} \; ({...}{:}t){:}r \; fb \; | \;
\mathbf{fun} \; (\vec{n{:}t}){:}r \; fb \; | \;
\mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
fb ::= & \;\; s \;;\; \mathbf{return} \; el\\
n ::= & \;\; {\it name}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}

@@

We use three different typing relations to present our typing rules
as a deduction system for these typing relations.
We use the relation $\env_{1}, \penv \vdash s, \env_{2}$ for typing
statements.
This relation means that given the type environment $\env_{1}$
and the projection environment $\penv$,
typing the statement $s$ produces the new type environment $\env_{2}$.
We use the relation $\env_{1}, \penv \vdash e : t, \env_{2}$ for
typing expressions.
This relation means that given the type environment $\env_{1}$
and the projection environment $\penv$,
the expression $e$ has type $t$ and produces the new type environment $\env_{2}$.
We use the relation $\env_{1}, \penv \vdash el : r_{1}, \env_{2}, (x,r_{2})$
for typing expression lists.
This relation means that given type type environment $\env_{1}$
and the projection environment $\penv$,
the expression list $el$ has type $r_{1}$ and produces the new type
environment $\env_{2}$ and produces the pair $(x,r_{2})$.
We need three typing relations because statements do not have types
and expression lists can result in a union of tuple types that
should be introduced in the projection environment.
We use $\env_{1}, \penv \vdash el : r, \env_{2}$ when the expression
list does not result in a union of tuple types.
We need two environments because $\env$ maps variables to types,
while $\penv$ maps projection variables to unions of tuple types.
We use $\env_{1}[n \mapsto t]$ for meaning that we extended the environment
$\env_{1}$ with the variable $n$ that maps to type $t$.
We use $\penv[x \mapsto r]$ for meaning that we extended the environment
$\penv$ with the projection variable $x$ that maps to the union of tuple types $r$.

Lua has multiple assignments, and the rule \textsc{T-ASSIGNMENT} uses
tuple types along with consistent-subtyping to type check this Lua feature:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\begin{array}{c}
\dfrac{\env_{1}, \penv \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
       r_{1} \lesssim r_{2}}
      {\env_{1}, \penv \vdash \vec{l} = el,\env_{3}}
\end{array}
\end{array}
\]

@@

First, this rule uses the same environment $\env$ to type check each
left-hand side value $l_{k}$, as they can perform different changes
in the environment.
Then, it takes each type $t_{k}$ and builds the tuple type for
the assignment rule.
It places the type $\Value{*}$ in the end to discard extra values.
The rule succeeds if the predicate \emph{merge} can
produce a new environment, which includes all the modifications
of each environment $\env_{k}$.
Intuitively, the predicate \emph{merge} fails when an environment
tries to change the type of a variable towards a type that is not
a subtype of its previous type.
We will show in this section that the refinement of table types can
change the environment, and that is the reason why we need to check
whether the environment is consistent.

@@

When our type system type checks an expression list,
it always include $\Nil{*}$ in the end of the type of this expression list,
if the type does not end in a variadic type.
Our type system has this behavior because $\Nil{*}$ along with subtyping
and consistent-subtyping use $\Nil$ to replace missing values.
This behavior preserves the semantics of Lua on replacing missing values,
as it is necessary when we omit optional parameters in a function call,
like the previous example showed.

@@

We do not allow the refinement of table types to change the type of
a field that is already present in the table type.
For instance,
\[
\string_ENV[``x"] \; {<}\String{>} = ``foo" \;;\; \string_ENV[``x"] \; {<}\Integer{>} = 1
\]
do not type check, as we are trying to add a field that already exists
in $\string_ENV$.
We could have used union types to allow the refinement of existing fields,
but this could lead to fields that have a too general type, which could
decrease the amount of static type checking.
This means that our type system does not allow the table type to change
towards a type that is not a subtype of the previous type.
In this example,
\[
\{``x":\String \cup \Integer\}_{open} \not\subtype \{``x":\String\}_{open}
\]

@@

Sometimes we want to use a local variable that holds an \emph{unique}
table type to initialize a \emph{closed} table type, as \emph{unique}
table types allow us to initialize optional fields.
To do that, Typed Lua includes a cast expression that allows us
to use an \emph{unique} table type before it becomes \emph{open}.
The rule \textsc{T-CAST} describes this behavior:
\[
\begin{array}{c}
\mylabel{T-CAST}\\
\dfrac{\env_{1}(n) \subtype t \;\;\;
       \env_{1}[n \mapsto t], \penv \vdash n_{e}:t_{1}, \env_{2}}
      {\env_{1}, \penv \vdash {<}t{>} \; n:t_{1}, \env_{2}}
\end{array}
\]

@@

The rule \textsc{T-FUNCTION3} uses the predicates \emph{closeall}
and \emph{closeset} to close \emph{open} and \emph{unique} table types.
The predicate \emph{closeall} closes all \emph{open} and \emph{unique}
table types before type checking another scope.
The predicate \emph{closeset} closes a given set of free assigned variables,
which is given by the predicate \emph{fav}.
This rule prevents the following unsafe example to type check:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; b:\Integer \times \Integer \times \Void \rightarrow \Integer \times \Void =$}\\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer, y:\Integer):\Integer$}\\
& & & \multicolumn{1}{l}{$a[``z"] \; {<}\Integer{>} = 1; \; \mathbf{return} \; x + y$}\\
& \multicolumn{3}{l}{$\mathbf{in} \; a[``z"] \; {<}\String{>} = ``foo"$}
\end{tabular}
\end{center}

Typed Lua also uses the refinement of table types to type check
common idioms that Lua programmers use for building modules and objects.
For instance, the rule \textsc{T-METHOD3} illustrates how our type system
type checks method definitions:
\[
\begin{array}{c}
\mylabel{T-METHOD3}\\
\dfrac{\begin{array}{c}
       \env_{1}(n_{1}) = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{unique|open} \;\;\;
       \env_{1}, \penv \vdash n_{2} : l \;\;\;
       \not \exists i \in 1..n \; l \lesssim k_{i}\\
       closeall(\env_{1}[self \mapsto \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{closed}, \vec{n} \mapsto \vec{t}, \ret \mapsto r]), \penv \vdash s, \env_{2}\\
       t_{m} = \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}, l{:}\Const \; \Self \times \vec{t} \rightarrow r\}_{unique|open}
       \end{array}}
      {\begin{array}{c}
       \env_{1}, \penv \vdash \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; s,\\
       closeset(\env_{1}[n_{1} \mapsto t_{m}, fav(\mathbf{fun} \; (\vec{n{:}t}){:}r \; s))
       \end{array}}
\end{array}
\]

The rule \textsc{T-METHOD3} allows our type system to type check the
following example, which is a translation of the definition of the class
\texttt{Shape} from Section \ref{sec:oop}:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{open} = \{ [``x"] = 0.0, [``y"] = 0.0 \}$}\\
\multicolumn{4}{l}{$\mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{fun} \; s{:}new (x:\Number, y:\Number):\Self \times \Void$}\\
& & \multicolumn{2}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{closed} =$}\\
& & & \multicolumn{1}{l}{$setmetatable(\{\}, \{ [``\string_\string_index"] = self \})$}\\
& & \multicolumn{2}{l}{$\mathbf{in} \; s[``x"] = x; \; s[``y"] = y; \; \mathbf{return} \; s$}\\
; & \multicolumn{3}{l}{$\mathbf{fun} \; s{:}move (x:\Number, y:\Number):\Void$}\\
& & \multicolumn{2}{l}{$self[``x"] = self[``x"] + x;$}\\
& & \multicolumn{2}{l}{$self[``y"] = self[``y"] + y$}
\end{tabular}
\end{center}

@@

Another common idiom that programmers use in Lua is to overload
the input parameter of functions, and use the function \texttt{type}
to execute different actions according to their types.
The rule \textsc{T-IF2} shows the case where our type system
discriminates the type of a local variable base on the tag \texttt{string}:
\[
\begin{array}{c}
\mylabel{T-IF2}\\
\dfrac{\begin{array}{c}
       \env_{1}(n) = t\\
       closeall(\env_{1}[n \mapsto \phi(t,\String)]), \penv \vdash s_{1}, \env_{2} \\
       closeall(\env_{1}[n \mapsto \phi(t,filter(t, \String))), \penv \vdash s_{2}, \env_{3}\\
       \env_{4} = closeset(\env_{1}[n \mapsto t], fav(s_{1}) \cup fav(s_{2}))
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; type(n) == ``string" \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}

@@

In Typed Lua, \emph{type} is a primitive that triggers a type change.
In the rule \textsc{T-IF2}, the type of the variable $n$ changes from
$t$ to $\phi(t,\String)$ inside the $\mathbf{if}$ branch, and it changes
the type of the variable $n$ from $t$ to $\phi(t,filter(t,\String))$
inside the $\mathbf{else}$ branch.
Our type system changes the type of a local variable to a filter type
because it gives the discriminated type when we use the variable as
an expression, and it restores the original type when we use the variable
in an assignment.
We will use the following example to discuss how our type system uses
the rule \textsc{T-IF2} to type check this given example:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; o:\String \times \String \cup \Integer \times \Void \rightarrow \String \times \Void =$}\\
& \multicolumn{3}{l}{$\mathbf{fun} \; (a:\String, b:\String \cup \Integer):\String \times \Void$}\\
& & \multicolumn{2}{l}{$\mathbf{local} \; r:\String = ``" \; \mathbf{in}$}\\
& & & \multicolumn{1}{l}{$\mathbf{if} \; type(b) == ``string" \; \mathbf{then} \; r = a \;{..}\;b \; \mathbf{else} \; r = rep(a, b)_{e}\; ;$}\\
& & & \multicolumn{1}{l}{$\mathbf{return} \; r$}\\
\multicolumn{4}{l}{$\mathbf{in} \; o(``foo", 2)_{s}$}
\end{tabular}
\end{center}

In this example, we are assuming that the function \emph{rep} is in
the environment and has type
$\String \times \Integer \times \String \cup \Nil \times \Void \rightarrow \String \times \Void$.
Note that rule \textsc{T-IF2} changes the type of $b$ to
$\phi(\String \cup \Integer, \String)$ inside the $\mathbf{if}$ branch,
and it changes the type of $b$ to
$\phi(\String \cup \Integer, \Integer)$ inside the $\mathbf{else}$ branch.
This example type checks because reading $b$ results in the type
$\String$ inside the $\mathbf{if}$ branch, and it results in the type
$\Integer$ inside the $\mathbf{else}$ branch.
Outside the condition, the local variable $b$ has type $\String \cup \Integer$.

Typed Lua also includes similar rules to handle the tags \texttt{nil},
\texttt{boolean}, and \texttt{number}.
There is also a similar rule for handling the tag \texttt{integer}, but
it is limited to Lua 5.3, as it depends on the function \texttt{math.type}.
This function appears only in Lua 5.3 and it returns the string $``integer"$
when its input parameter is a number that has an integer representation,
the string $``float"$ when its input parameter is a number that has a
floating point representation, or $\Nil$ otherwise.

Lua programmers also overload the return type of functions to denote errors,
and our type system also handles this idiom.
For instance,
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; q:\pi_{1}^{x}, r:\pi_{2}^{x} = idiv(1, 2)_{m}^{x} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; print(q + r)_{s} \; \mathbf{else} \; print(``ERROR: " \; .. \; r)_{s}$}
\end{tabular}
\end{center}
type checks, assuming that functions \emph{idiv} and \emph{print} are in
the environment with the respective types
$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void$
and
$\Value{*} \rightarrow \Void$.
There are three typing rules involved in the type checking of this example,
which we will explain now.

First, type checking $idiv(1, 2)_{m}^{x}$ results in a tuple of projection
types and in a pair that the typing rule of the local declaration needs
for introducing the variable $x$ in the projection environment.
The rule \textsc{T-EXPLIST6} express this intuition:
\[
\begin{array}{c}
\mylabel{T-EXPLIST6}\\
\dfrac{\begin{array}{c}
       \env, \penv \vdash me^{x}:r\\
       r = t_{1} \times ... \times t_{n} \times \Void \sqcup t_{1}' \times ... \times t_{n}' \times \Void, \env_{1}
       \end{array}}
      {\env, \penv \vdash me^{x}:\pi_{1}^{x} \times ... \times \pi_{n}^{x} \times \Nil{*}, \env_{1}, (x,r)}
\end{array}
\]

Then, type checking the local declaration checks whether the tuple of projection
types match the types in the annotations.
If the types match, the local declaration extends the type environment with the
new variables, and it also uses the pair $(x,r)$ from the previous rule to
extend the projection environment before type checking the statements.
The rule \textsc{T-LOCAL} express this intuition:
\[
\begin{array}{c}
\mylabel{T-LOCAL}\\
\dfrac{\env_{1}, \penv \vdash el:r_{1}, \env_{2}, (x,r_{2}) \;\;\;
       r_{1} \lesssim \vec{t} \;\;\;
       \env_{2}[\vec{n} \mapsto \vec{t}], \penv[x \mapsto r_{2}] \vdash s, \env_{3}}
      {\env_{1}, \penv \vdash \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s, \env_{3} - \{\vec{n} \mapsto \vec{t}\}}
\end{array}
\]

After type checking the statement $s$, the rule \textsc{T-LOCAL} produces a
new environment without the variables that it introduced before type checking $s$.

Once the projection is in the environment, our type system can allow
discriminating variables that hold projection types.
The rule \textsc{T-IF3} illustrates the case of our previous example:
\[
\begin{array}{c}
\mylabel{T-IF3}\\
\dfrac{\begin{array}{c}
       \env_{1}(n) = \pi_{i}^{x} \;\;\; \penv(x) = r\\
       closeall(\env_{1}), \penv[x \mapsto fpt(r,\Nil,i)] \vdash s_{1}, \env_{2} \\
       closeall(\env_{1}), \penv[x \mapsto gpt(r,\Nil,i)] \vdash s_{2}, \env_{3}
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; n \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2}))}
\end{array}
\]

The rule \textsc{T-IF3} uses the predicates \emph{fpt} and \emph{gpt}
to filter projections, affecting all variables that bind to the same projection.
For instance, our previous example type checks through the rule \textsc{T-IF3},
because it makes the rule \textsc{T-IF3} use the predicate
\[
fpt(\Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void, \Nil, 1)
\]
to discriminate the projection $x$ to the single tuple
$\Integer \times \Integer \times \Void$, and the predicate
\[
gpt(\Integer \times \Integer \times \Void \sqcup \Nil \times \String \times \Void, \Nil, 1)
\]
to discriminate the projection $x$ to the single tuple
$\Nil \times \String \times \Void$.
This means that the type of $q$ ($\pi_{1}^{x}$) and $r$ ($\pi_{2}^{x}$)
project to $\Integer$ and $\Integer$ inside the $\mathbf{if}$ branch,
but they project to $\Nil$ and $\String$ inside the $\mathbf{else}$ branch.
Outside the $\mathbf{if}$ statement, the types of $q$ and $r$ project
to $\Integer \cup \Nil$ and $\Integer \cup \String$, respectively.
